Beating Simple Backtracking Search in Subgraph Algorithms\thanks{This work was supported by the Engineering and Physical Sciences Research Council [grant numbers EP/026842/1 and EP/M508056/1].}

### 2018-11-29 paper/paper.tex
Last tweaks. Definitely this time. Honest

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. The problem is broadly applicable, with applications in areas
including bioinformatics \cite{DBLP:journals/bmcbi/BonniciGPSF13}, chemistry \cite{o:Regin95}, computer vision
\cite{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, law enforcement
\cite{DBLP:journals/cacm/CoffmanGM04}, model checking \cite{DBLP:journals/tcs/SevegnaniC15},
malware detection \cite{DBLP:conf/dimva/BruschiMM06}, compilers
\cite{DBLP:conf/cgo/MurrayF12,DBLP:conf/cp/BlindellLCS15}, pattern recognition
\cite{DBLP:journals/ijprai/FoggiaPV14}, program similarity
comparison \cite{o:DallaPredaV17}, the design of mechanical locks
\cite{DBLP:journals/siamrev/VomelLBF17}, and graph databases
\cite{DBLP:journals/jair/McCreeshPST18}.

@@

Our experiments are performed on the EPCC Cirrus HPC facility, on systems with dual Intel Xeon
E5-2695 v4 CPUs and 256GBytes RAM, running Centos 7.3.1611. We use GCC 7.2.0 as the compiler. For
paralellism, we use C++ native threads, and for distributed parallelism we also use the SGI MPT
implementation of MPI. All timing measurements are steady-clock, and we use a deterministic
pseudo-random number generator for reproducibility.

@@

We begin with a set of baseline performance measurements. In the top
two plots of \cref{figure:others} we show the cumulative number of instances solved over time for the
non-induced and induced problems respectively. We compare the Glasgow Subgraph Solver using
depth-first search (DFS) and with the modifications described in the remainder of this paper
(solution-biased search, SBS), the PathLAD variation of the LAD algorithm
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, VF2
\cite{DBLP:journals/pami/CordellaFSV04}, RI \cite{DBLP:journals/bmcbi/BonniciGPSF13}, and VF3
\cite{DBLP:conf/gbrpr/CarlettiFSV17} (which only supports the induced problem), in each case using
the original implementation provided by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).  The plots show that our starting point
comfortably beats PathLAD, VF2, VF3 and RI, except for very low choices of timeout.

The dataset includes many instances which are extremely easy for a good solver, and so it can be
hard to see the differences between the stronger solvers at higher runtimes. This paper focusses
upon improving the performance on the remaining hard satisfiable instances, and so in the bottom two
plots in \cref{figure:others} (and in subsequent cumulative plots) we show only satisfiable
instances, and use a reduced range on both axes.  For the remainder of this paper, we show only the
non-induced problem, which tends to be harder; results with the induced variant are similar.

@@

A \emph{discrepancy} is where search goes against the advice of a value-ordering heuristic.
Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97,DBLP:conf/cpaior/KarouiHLN07}
are alternatives to backtracking search that initially search disallowing all discrepancies, and
then retry search allowing an increasing number of discrepancies at each iteration until either a
solution is found or unsatisfiability is proven. These schemes assume that value-ordering heuristics
are usually reliable, and that most solutions can be found with only a small number of
discrepancies. In such cases, the heavy commitment to early branching choices made by conventional
backtracking search can be extremely costly.

@@

In the more detailed view in the right-hand plot of \cref{figure:old-vs-new}, comparing the original
algorithm to solution-biased search, all of the unsatisfiable instances are very close to the $x-y$
diagonal, showing that their performance is nearly unchanged. On the other hand, there are large
numbers of satisfiable instances well below the diagonal line, indicating large speedups.  Better
yet, there are only a handful of satisfiable instances that are more than a factor of ten times
worse.  In other words, as well as improving performance, we have made up most of the consistency we
lost by introducing randomness.  This is a promising result: for the first time, we have introduced
a viable alternative to backtracking search in a setting where we both have a strong
\emph{value}-ordering heuristic, \emph{and} care about both satisfiable and unsatisfiable instances.

@@

To estimate solution density, we need an estimate of how big different subproblems are likely to be,
and of how many solutions each subproblem is likely to contain. Of course, obtaining exact (or even
approximate) values for these figures is at least as hard as solving the problem in its entirety.
For pairs of Erd\H{o}s-R\'enyi
random graphs with large solution counts (i.e.\ chosen from within the ``easy satisfiable'' region
\cite{DBLP:journals/jair/McCreeshPST18}), we observe a linear relationship between subproblem size
and number of solutions. Thus, for graphs from this distribution, we need only an estimator of
subproblem size. Measurements also suggest that, for pairs of Erd\H{o}s-R\'enyi graphs, subproblems
under a target vertex of degree $d$ tend to contain a small constant times more search nodes than
those under a target vertex of degree $d - 1$.  This empirical analysis suggests that an estimator
that is exponential in $d$ will give our method the desired behaviour, at least for
Erd\H{o}s-R\'enyi graphs. We expect it may be possible to derive better estimators for particular
input classes, although over the full range of problem instances, we have verified that exponential
estimators substantially outperform polynomial and factorial weightings.

@@

    \caption{Above, the cumulative number of instances solved over time, comparing the sequential
    algorithm to results using 32 threads on a single machine, and using five or ten distributed
    memory hosts. Below, instance by instance comparisons.}\label{figure:parallel}
\end{figure}

@@

It is important to emphasise that there is no expectation of a linear speedup from this approach,
particularly for satisfiable instances. Due to the biased random nature of the value-ordering heuristic,
one of our extra threads may ``get lucky'' and find a solution much quicker than we would
sequentially, leading to a superlinear speedup. Conversely, our extra threads may contribute nothing to finding a
solution---or worse, due to new nogoods altering the choice made by the random branching heuristic,
we could even get an absolute slowdown. We can see both of these effects in the bottom left plot of
\cref{figure:parallel}: although we achieve roughly a linear speedup on unsatisfiable instances,
satisfiable instances show much greater variability. However, this approach does at least mirror our intuition of
allocating search effort in proportion to where the value-ordering heuristic believes it will be
most fruitful, and so we should not be too surprised that we see roughly a linear speedup on average
on harder instances.

@@

\Cref{figure:parallel} also shows the results of these experiments, using five, ten, and twenty
hosts.  Because each host has two sockets, the five host results use ten MPI ranks, each with
eighteen threads, and the ten and twenty host results use twenty and forty MPI ranks respectively.
The supercomputing service we use is not designed for huge numbers of very short problems, and so we
ran only instances whose sequential runtime was at least one second; for the sake of plotting
results, we treat skipped instances as taking one second. Due to the job launcher used, it is also
not possible to accurately measure ``total'' runtime including startup costs, and so instead we
report the runtime of the rank zero process---this figure is somewhat optimistic for very easy
instances. With these caveats in mind, when seeing how long a timeout is needed to solve any 14,370
instances, we get aggregate speedups of 57.1 from five hosts and 95.5 from ten hosts over a
sequential baseline; using twenty hosts is slightly slower, due to increased overheads. However,
when seeing how long a timeout is needed to solve 14,424 instances, distributed parallelism gives us
further aggregate speedups of 2.2, 3.5, and 7.9 from five, ten, and twenty hosts, over the single
machine parallel baseline.

@@

The conventional view of value-ordering heuristics is that they define a search order. We have
proposed an alternative perspective, where value-orderings define a weighting specifying how much
search effort should be put into different subproblems, based upon a rough estimate of solution
densities. We have also shown how to turn this perspective into an algorithm, by combining a biased
random value-ordering heuristic with rapid restarts and nogood recording. This combination of
techniques gives us, for the first time, a practical alternative to backtracking search where we
have a strong value-ordering heuristic and where we care both about satisfiable and unsatisfiable
instances.  A further benefit is the ease with which such a search can be parallelised. By having
each thread carry out the same search with a different random seed, and sharing nogoods only on
restarts, we remove the need for intrusive changes to the core search algorithm, eliminate the
irregularity problem, and still respect the advice of the value-ordering heuristic.

>>>>

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. The problem is broadly applicable, arising in areas
including bioinformatics \cite{DBLP:journals/bmcbi/BonniciGPSF13}, chemistry \cite{o:Regin95}, computer vision
\cite{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, law enforcement
\cite{DBLP:journals/cacm/CoffmanGM04}, model checking \cite{DBLP:journals/tcs/SevegnaniC15},
malware detection \cite{DBLP:conf/dimva/BruschiMM06}, compilers
\cite{DBLP:conf/cgo/MurrayF12,DBLP:conf/cp/BlindellLCS15}, pattern recognition
\cite{DBLP:journals/ijprai/FoggiaPV14}, program similarity
comparison \cite{o:DallaPredaV17}, the design of mechanical locks
\cite{DBLP:journals/siamrev/VomelLBF17}, and graph databases
\cite{DBLP:journals/jair/McCreeshPST18}.

@@

Our experiments are performed on the EPCC Cirrus HPC facility, on systems with dual Intel Xeon
E5-2695 v4 CPUs and 256GBytes RAM, running Centos 7.3.1611. We use GCC 7.2.0 as the compiler. For
parallelism, we use C++ native threads, and for distributed parallelism we also use the SGI MPT
implementation of MPI. All timing measurements are steady-clock, and we use a deterministic
pseudo-random number generator for reproducibility.

@@

We begin with a set of baseline performance measurements. In the top
two plots of \cref{figure:others} we show the cumulative number of instances solved over time for the
non-induced and induced problems respectively. We compare the Glasgow Subgraph Solver using
depth-first search (DFS) and with the modifications described in the remainder of this paper
(solution-biased search, SBS), the PathLAD variation of the LAD algorithm
\cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, VF2
\cite{DBLP:journals/pami/CordellaFSV04}, RI \cite{DBLP:journals/bmcbi/BonniciGPSF13}, and VF3
\cite{DBLP:conf/gbrpr/CarlettiFSV17} (which only supports the induced problem), in each case using
the original implementation provided by the algorithm's authors.  The plots show that our starting
point comfortably beats PathLAD, VF2, VF3 and RI, except for very low choices of timeout. For each
algorithm, the $y$ value gives the cumulative number of instances which (individually) can be solved
in no more than $x$ milliseconds.  The vertical distance between two lines therefore shows how many
more instances can be solved by one solver than another, if every instance is run separately with
the chosen $x$ timeout. The horizontal distance shows how many times longer the per-instance timeout
would need to be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances
(bearing in mind that the two sets of $y$ instances could be different), and gives a measure called
\emph{aggregate speedup} \cite{DBLP:conf/cpaior/HoffmannMNPRS018}.

The dataset includes many instances which are extremely easy for a good solver, and so it can be
hard to see the differences between the stronger solvers at higher runtimes. This paper focusses
upon improving the performance on the remaining hard satisfiable instances, and so in the bottom two
plots in \cref{figure:others} (and in subsequent cumulative plots for sequential algorithms) we show
only satisfiable instances, and use a reduced range on both axes.  For the remainder of this paper,
we show only the non-induced problem, which tends to be harder; results with the induced variant are
similar.

@@

A \emph{discrepancy} is where search goes against the advice of a value-ordering heuristic.
Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97,DBLP:conf/cpaior/KarouiHLN07}
are alternatives to backtracking search that initially search disallowing all discrepancies, and
then retry search allowing an increasing number of discrepancies at each iteration until either a
solution is found or unsatisfiability is proven. These schemes assume that value-ordering heuristics
are usually reliable, and that most solutions can be found with only a small number of
discrepancies. In such cases, the heavy commitment to early branching choices made by
backtracking search can be extremely costly.

@@

In the more detailed view in the right-hand plot of \cref{figure:old-vs-new}, comparing the original
algorithm to solution-biased search, all of the unsatisfiable instances are very close to the $x-y$
diagonal, showing that their performance is nearly unchanged. On the other hand, there are large
numbers of satisfiable instances well below the diagonal line, indicating large speedups.  Better
yet, there are only a handful of satisfiable instances that are more than a factor of ten times
worse.  In other words, as well as improving performance, we have made up most of the consistency we
lost by introducing randomness.

@@

To estimate solution density, we need an estimate of how big different subproblems are likely to be,
and of how many solutions each subproblem is likely to contain. Of course, obtaining exact (or even
approximate) values for these figures is at least as hard as solving the problem in its entirety,
but we may obtain usable approximations.  For pairs of Erd\H{o}s-R\'enyi
random graphs with large solution counts (i.e.\ chosen from within the ``easy satisfiable'' region
\cite{DBLP:journals/jair/McCreeshPST18}), we can observe a linear relationship between subproblem size
and number of solutions. Thus, for graphs from this distribution, we need only an estimator of
subproblem size. Measurements also suggest that, for pairs of Erd\H{o}s-R\'enyi graphs, subproblems
under a target vertex of degree $d$ tend to contain a small constant times more search nodes than
those under a target vertex of degree $d - 1$.  This empirical analysis suggests that an estimator
that is exponential in $d$ will give our method the desired behaviour, at least for
Erd\H{o}s-R\'enyi graphs. We expect it may be possible to derive better estimators for particular
input classes, although over the full range of problem instances, we have verified that exponential
estimators substantially outperform polynomial and factorial weightings.

@@

    \caption{Above, the cumulative number of instances solved over time, comparing the sequential
    algorithm to results using 32 threads on a single machine, and using five, ten or twenty
    distributed memory hosts. Below, instance by instance comparisons.}\label{figure:parallel}
\end{figure}

@@

It is important to emphasise that there is no expectation of a linear speedup from this approach
\cite{DBLP:conf/aaai/LiW84,DBLP:journals/tc/LiW86,DBLP:conf/irregular/BruinKT95},
particularly for satisfiable instances. Due to the biased random nature of the value-ordering heuristic,
one of our extra threads may ``get lucky'' and find a solution much quicker than we would
sequentially, leading to a superlinear speedup. Conversely, our extra threads may contribute nothing to finding a
solution---or worse, due to new nogoods altering the choice made by the random branching heuristic,
we could even get an absolute slowdown. We can see both of these effects in the bottom left plot of
\cref{figure:parallel}: although we achieve roughly a linear speedup on unsatisfiable instances,
satisfiable instances show much greater variability. However, this approach does at least mirror our intuition of
allocating search effort in proportion to where the value-ordering heuristic believes it will be
most fruitful, and so we should not be too surprised that we see roughly a linear speedup on average
on harder instances.

@@

\Cref{figure:parallel} also shows the results of these experiments, using five, ten, and twenty
hosts.  Because each host has two CPU sockets, the five host results use ten MPI ranks, each with
eighteen threads, and the ten and twenty host results use twenty and forty MPI ranks respectively.
The supercomputing service we use is not designed for huge numbers of very short problems, and so we
ran only instances whose sequential runtime was at least one second; for the sake of plotting
results, we treat skipped instances as taking one second. Due to the job launcher used, it is also
not possible to accurately measure ``total'' runtime including startup costs, and so instead we
report the runtime of the rank zero process---this figure is somewhat optimistic for very easy
instances. With these caveats in mind, when seeing how long a timeout is needed to solve any 14,370
instances, we get aggregate speedups of 57.1 from five hosts and 95.5 from ten hosts over a
sequential baseline; using twenty hosts is slightly slower, due to increased overheads. However,
when seeing how long a timeout is needed to solve 14,424 instances, distributed parallelism gives us
further aggregate speedups of 2.2, 3.5, and 7.9 from five, ten, and twenty hosts, over the single
machine parallel baseline.

@@

The conventional view of value-ordering heuristics is that they define a search order. We have
proposed an alternative perspective, where value-orderings define a weighting specifying how much
search effort should be put into different subproblems, based upon a rough estimate of solution
densities. We have also shown how to turn this perspective into an algorithm, by combining a biased
random value-ordering heuristic with rapid restarts and nogood recording. This combination of
techniques gives us, for the first time, a practical alternative to backtracking search where we
have a strong \emph{value}-ordering heuristic, and where we care both about satisfiable \emph{and}
unsatisfiable instances.

A further benefit is the ease with which such a search can be parallelised. By having each thread
carry out the same search with a different random seed, and sharing nogoods only on restarts, we
remove the need for intrusive changes to the core search algorithm, eliminate the irregularity
problem, and still respect the advice of the value-ordering heuristic.
